---
title:  "Basic Goodness-of-fit Diagnositic Plots"
author: "Certara PMxO R Tools Workstream"
date:   "`r format(Sys.Date(), '%d-%b-%Y')`"
output: rmarkdown::html_vignette
    
vignette: >
  %\VignetteIndexEntry{Basic Goodness-of-fit Diagnositic Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
    font-family: "Arial", Arial, sans-serif;
    font-size: 10pt;
    border-collapse: collapse;
    padding: 0px;
    margin: 0px;
    width: 100%;
    display: block;
    overflow: auto;
}
</style>

```{r echo=FALSE}
knitr::opts_knit$set(root.dir=system.file("sample-data", package="ggcertara"))
knitr::opts_chunk$set(warning=FALSE, message=TRUE, comment="#>", dpi=300, out.width="80%")
```


# Setup

Before starting, load the `ggplot2` and `ggcertara` packages, and set the
default `ggplot` theme to `theme_certara()`:

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggcertara)
})
theme_set(theme_certara())
```

# Basic usage

In the most basic case, you simply need to call the function `gof()` from
within the directory that contains a model output table. For this example,
we don't have an actual model so we will use some sample data that has been
provided with this packaga and pretend that it is the  result of a model
fit. We check the current working directory and note that it contains a
single CSV file.

```{r}
getwd()
dir()
```

This file was actually generated by NONMEM (using the statement `$TABLE
FILE=nmtable1.csv FORMAT=,1PE15.8 ...`).  It's contents look like this:

```{r echo=F}
knitr::kable(head(nmtable1))
```


When called without a `data` argument, the `gof()` function will actually
search the current directory for a CSV file that looks like it contains the
information it requires (i.e., has columns `dv`, `pred`, `ipred` and so on;
the data will also be filtered on an `mdv` column, if present).
Alternatively, the `sdtab` file that is commonly used with PsN/Xpose will
also be detected.

```{r fig.width=6, fig.height=9, out.width="70%"}
gof()
```

By default, the `gof()` function produces a figure with 6 panels arranged in
a 3-by-2 grid: DV vs. IPRED, DV vs. PRED, CWRES vs. PRED, CWRES vs. TIME,
CWRES vs. TAD, and absolute value of IWRES vs. IPRED. Note that all axes are
nicely labelled (though they lack units at this point). With this simple
function call, we have produced a report ready figure!

# Customizing the output

## Data specification

While it is convenient to have the function automatically scan for data in
the current working directory, this may not always be the desired approach
and it is also possible to pass a standard `data.frame` to the `gof()`
function via the `data` argument.

The `gof()` function expects the dataset to contain standard column names:

  - time:   Time (e.g., time since first dose)
  - tad:    Time after dose
  - dv:     Dependent variable
  - pred:   Population prediction
  - ipred:  Individual prediction
  - cwres:  Conditional weighted residual
  - iwres:  Individual weighted residual

Not all of these need to be present, only those that are needed for the
disired plots (for instance, if no plots involving IWRES are requested, then
it does not need to be present). The column names must be in lower case.

We now read in the sample data used above and for the remainder of this
vignette we will pass this data to the `gof()` function. We remember to
convert the column names to lower case and to filter out only the rows that
contain observations (as indicated by MDV=0).

```{r}
dat <- read.csv("nmtable1.csv")
names(dat) <- tolower(names(dat))
dat <- subset(dat, mdv==0)
```

## Selecting panels

The `gof()` function can easily produce a variety of standard panels. The
panels are numbered:

  1.  Histogram of CWRES
  2.  QQ-plot of CWRES
  3.  DV vs. IPRED linear scale
  4.  DV vs. PRED linear scale
  5.  DV vs. IPRED log scale
  6.  DV vs. PRED log scale
  7.  CWRES vs. PRED
  8.  CWRES vs. TIME
  9.  CWRES vs. TAD
  10. |IWRES| vs. IPRED
  11. |IWRES| vs. TIME
  12. |IWRES| vs. TAD

Using the `panels` argument, single or multiple panels can be selected.  For
example, to prduce both a histogram and QQ-plot of CWRES, select panels 1
and 2:

```{r fig.width=6, fig.height=3}
gof(dat, panels=1:2)
```

To obtain just a plot of DV vs. IPRED on log-log scale, select panel 5:

```{r fig.width=3, fig.height=3, out.width="50%"}
gof(dat, panels=5)
```

### The `gof_list` object

Individual panels can also be obtain in 2 other ways.
The first is to select the desired panel from the list of panels returned by
`gof_list()` (which returns a list of all standard panels):

```{r fig.width=3, fig.height=3, out.width="50%"}
p <- gof_list(dat)
p[[5]]
```

The other way is to use a named function:

```{r fig.width=3, fig.height=3, out.width="50%"}
gof_dv_vs_ipred(dat, log_xy=T)
```

Note that individual panels are standard `ggplot` objects (that can be
further customized, modified, annotated, composed, etc.), while when
multiple panels are requested a `patchwork` object is returned (this can
also be further cutomized, see below).

## Labels

```{r}
getOption("gof.labels")
```

## Outliers

To highlight outliers using a special symbol (and color), you first define a
column in the dataset which is a factor with 2 levels (be mindful of the
labels, as these will appear in the legend):

```{r}
dat$outlier <- factor(abs(dat$cwres) > 2.5, labels=c("|CWRES| \U{2264} 2.5", "|CWRES| > 2.5"))
```

Then, use th `highlight` option, like this:

```{r fig.width=6, fig.height=10, out.width="70%"}
gof(dat, highlight=outlier)
```

This feature should be used sparingly.

## Changing aesthetics (colors, etc.)

For a single panel, you can just modify the `ggplot` object:

```{r fig.width=4, fig.height=4, out.width="60%"}
gof(dat, panels=3, highlight=outlier) +
  scale_colour_manual(values=c("#7059A6", "#52ccbb")) +
  scale_shape_manual(values=c(19, 19)) +
  scale_size_manual(values=c(1, 5))
```

For a `patchwork` object (multiple panels), you can do the same thing by
replacing the `+` operator with a `%`:


```{r fig.width=6, fig.height=4}
gof(dat, panels=3:4, highlight=outlier) &
  scale_colour_manual(values=c("#7059A6", "#52ccbb")) &
  scale_shape_manual(values=c(19, 19)) &
  scale_size_manual(values=c(1, 5))
```

`patchwork` is smart enough to produce a single legend for both panels.

## Adding geoms and annotations

```{r fig.width=6, fig.height=3}
gof(dat, panels=10:11) &
  geom_smooth(method="lm", color="#52ccbb", se=F, size=1)
```


```{r fig.width=6, fig.height=3}
 gof(dat, panels=5:6) &
  geom_hline(yintercept=1.2, linetype="dashed") &
  geom_text(data=data.frame(x=Inf, y=1.2), aes(x=x, y=y),
    label="LLOQ = 1.2 ng/mL", hjust=1.1, vjust=-0.5, size=3)
```

# R session information

```{r}
sessionInfo()
```

```{r echo=F}
# vim: ts=2 sw=2 et

