---
title:  "Basic Goodness-of-fit Diagnostic Plots"
author: "Certara PMxO R Tools Workstream"
date:   "`r format(Sys.Date(), '%d-%b-%Y')`"
output: rmarkdown::html_vignette
    
vignette: >
  %\VignetteIndexEntry{Basic Goodness-of-fit Diagnostic Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
    font-family: "Arial", Arial, sans-serif;
    font-size: 10pt;
    border-collapse: collapse;
    padding: 0px;
    margin: 0px;
    width: 100%;
    display: block;
    overflow: auto;
}
</style>

```{css echo=FALSE}
/* Define a margin before h1 element */
h1  {
  margin-top: 3em;
}
h2  {
  margin-top: 2em;
}
h3  {
  margin-top: 1em;
}
``` 

```{r echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=TRUE, comment="#>", dpi=300, out.width="80%")
```


# Setup

Before starting, load the `ggplot2` and `ggcertara` packages, and set the default `ggplot` theme to `theme_certara()`:

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggcertara)
  library(kableExtra) #used for the vignette not for ggcertara demonstration
})
theme_set(theme_certara())
```

# Overview

The `ggcertara` package and `theme_certara` theme are meant to provide a standardized look to typical `gof` plots employed by pharmacometricians. The syntax is based on `ggplot2`. The `ggcertara` also utilizes the `patchwork` package and syntax. These will be touched upon later in this vignette.

# Data:

## Data Expectations
The `gof()` function uses standard column names in lower case. The following column names are recognized by the functions in this package:
  
  - time:   Time (e.g., time since first dose)
  - tad:    Time after dose
  - dv:     Dependent variable
  - pred:   Population prediction
  - ipred:  Individual prediction
  - cwres:  Conditional weighted residual
  - iwres:  Individual weighted residual

Not all of these need to be present, only those that are needed for the disired plots (for instance, if no plots involving `iwres` are requested, then it does not need to be present). The column names must be in lower case.

## Reading: 

The `gof_read_data()` function simplifies the data loading process by:

1. loading the data
2. converting all names to lower case
3. checking for the presence of an `mdv` column and, if present, filtering the rows to keep only those with `mdv = 0` 

For this vignette we will be using some sample data that is provided with the package.  The data is contained in a file named `nmtable.csv`, that was actually generated by NONMEM using a `$TABLE` statement (i.e., `$TABLE FILE=nmtable.csv FORMAT=,1PE15.8 ...`).  It's contents look like this:

```{r echo=F}
knitr::kable(head(nmtable))
```

In a normal project workflow, the location of the appropriate file would be
clear. For the vignette, the file can be located as follows:

```{r}
file <- system.file(package="ggcertara", "sample-data/nmtable.csv")
```

Now we load the file:

```{r}
dat <- gof_read_data(file)
```

Examining the data, we can see that the names are lower case and the rows have been filtered such that `mdv = 0`:

```{r}
head(dat)
table(dat$mdv)
```

# Basic usage

The basic functionality is accessed through the `gof()` function:

```{r fig.width=9, fig.height=6, out.width="90%"}
gof(dat) #Note: output aspect ratio is set to 3:2, in this case width=9, height=6
```

By default, the `gof()` function produces a figure with 6 panels arranged in
a 2-by-3 grid. 


```{r echo=FALSE, fig.width=6, fig.height=9, out.width="70%"}
dftemp <- data.frame(Column_1=c('DV vs. IPRED', 'DV vs. PRED'),
                 Column_2=c('CWRES vs. TIME','CWRES vs. PRED'),
                 Column_3=c('CWRES vs. TAD', '|IWRES| vs. IPRED'))
knitr::kable(dftemp,align = 'c') %>%
  column_spec(1, width = "15em", background = "white", include_thead = TRUE) %>%
  column_spec(2, width = "15em", background = "white", include_thead = TRUE) %>%
  column_spec(3, width = "15em", background = "white", include_thead = TRUE)
```

<div style="margin-bottom:20px;">
</div>

Note that all axes are nicely labelled (though they lack units at this point). With this simple function call, we have produced a report ready figure!

# Customizing the output

## Selecting panels

The `gof()` function can easily produce a variety of panels which can be referenced as either numbered panels,  named functions, or as strings.

```{r echo=FALSE}
dftemp <- data.frame(Number=c('1','2','3','4','5','6','7','8','9','10','11, -3','12, -4','13, -5','14, -6','15, -7','16, -8','17, -9','18, -10'),
                 'Function'=c('gof_cwres_histogram()','gof_cwres_qqplot()','gof_dv_vs_ipred()','gof_dv_vs_pred()','gof_cwres_vs_pred()','gof_cwres_vs_time()','gof_cwres_vs_tad()','gof_absiwres_vs_ipred()','gof_absiwres_vs_time()','gof_absiwres_vs_tad()','gof_dv_vs_ipred(log_xy=T)','gof_dv_vs_pred(log_xy=T)','gof_cwres_vs_pred(log_x=T)','gof_cwres_vs_time(log_x=T)','gof_cwres_vs_tad(log_x=T)','gof_absiwres_vs_ipred(log_x=T)','gof_absiwres_vs_time(log_x=T)','gof_absiwres_vs_tad(log_x=T)'),
                 Name=c('cwres_histogram','cwres_qqplot','dv_vs_ipred','dv_vs_pred','cwres_vs_pred','cwres_vs_time','cwres_vs_tad','absiwres_vs_ipred','absiwres_vs_time','absiwres_vs_tad','dv_vs_ipred_log','dv_vs_pred_log','cwres_vs_pred_log','cwres_vs_time_log','cwres_vs_tad_log','absiwres_vs_ipred_log','absiwres_vs_time_log','absiwres_vs_tad_log'),
                 Description=c('Histogram of CWRES','QQ-plot of CWRES','DV vs. IPRED','DV vs. PRED','CWRES vs. PRED','CWRES vs. TIME','CWRES vs. TAD','|IWRES| vs. IPRED','|IWRES| vs. TIME','|IWRES| vs. TAD','DV vs. IPRED log scale','DV vs. PRED log scale','CWRES vs. PRED log scale','CWRES vs. TIME log scale','CWRES vs. TAD log scale','|IWRES| vs. IPRED log scale','|IWRES| vs. TIME log scale','|IWRES| vs. TAD log scale'))
knitr::kable(dftemp,align = 'l') %>%
  column_spec(1, width = "5em", background = "white", include_thead = TRUE) %>%
  column_spec(2, width = "15em", background = "white", include_thead = TRUE) %>%
  column_spec(3, width = "15em", background = "white", include_thead = TRUE) %>%
  column_spec(4, width = "15em", background = "white", include_thead = TRUE)
```

Note that the standard 6 plots in the `gof()` function correspond to the numbered panels 3, 6, 7, 4, 5, 8 from this list.

<div style="margin-bottom:40px;">
</div>

### Calling the `panels` option within `gof()`
Using the `panels` argument, select panels can be selected using either the respective panel numbers or names (as strings).  For example, to prduce both a histogram and QQ-plot of CWRES, select panels 1 and 2 or callthem by name `cwres_histogram` and `cwres_qqplot`:

```{r fig.width=6, fig.height=3, out.width="60%"}
gof(dat, panels=1:2)
gof(dat, panels=c('cwres_histogram', 'cwres_qqplot'))
```

<div style="margin-bottom:40px;">
</div>

For log-scale plots (panels 11-18 above), we can call their panel number or add a `-` to their linear scale counterpart. For example, to obtain a plot of DV vs. IPRED on log-log scale, we can set `panel=11` or `panel=-3`:

```{r fig.width=6, fig.height=3, out.width="60%"}
gof(dat, panels=c(11,-3))
```

<div style="margin-bottom:40px;">
</div>

We can use the `layout` option to specify the number of rows and columns in a multi-panel output.

```{r fig.width=6, fig.height=9, out.width="60%"}
gof(dat, panels=c(3, -3, 4, -4, 5, 6), layout = c(3, 2))
```

<div style="margin-bottom:20px;">
</div>


### Named function:

We can also call specific panels using the corresponding named function. Log variants of the plot can be obtained by specifying either `log_x=T` or `log_xy=T` depending on the function.

```{r fig.width=3, fig.height=3, out.width="30%"}
gof_dv_vs_ipred(dat)
gof_dv_vs_ipred(dat, log_xy=T)
```

<div style="margin-bottom:40px;">
</div>

### The `gof_list` object

Individual panels can also be with the `gof_list()` function. This returns a numbered list of the 6 standard panels by default, or all panels if `all=T` is specified:

```{r}
p <- gof_list(dat, all=T)
p
```

<div style="margin-bottom:40px;">
</div>

Individual panels can be extracted from the list:

```{r fig.width=3, fig.height=3, out.width="30%"}
p[[5]]
```

<div style="margin-bottom:40px;">
</div>

`gof_list` can also be used to build a new list of plots, by specifying `empty=T`:

```{r fig.width=3, fig.height=3, out.width="30%"}
p <- gof_list(empty = T)
p[[1]] = gof_dv_vs_ipred(dat)
p[[2]] = gof_dv_vs_pred(dat)
```

<div style="margin-bottom:40px;">
</div>

A composite plot can be obtained from a list using the `gof_layout()` function:

```{r fig.width=6, fig.height=3, out.width="60%"}
gof_layout(p)
```


## Labels and units

There are several ways to customize the plot axis labels/units. When a recognized variable name is used in a plot (e.g. `ipred`, `cwres`, etc.), the functions will find a corresponding axis label in a list object which by default corresponds to a global option:

```{r echo=F}
dftemp <- data.frame(Option=c('gof.label.dv','gof.label.pred','gof.label.ipred','gof.label.cwres','gof.label.iwres','gof.label.time','gof.label.tad','gof.units.dv','gof.units.time','gof.units.tad'),
                 Default=c('Observed Value','Population Prediction','Individual Prediction','Conditional Weighted Residual','Individual Weighted Residual','Time','Time After Dose','NA','NA','NA'))

knitr::kable(dftemp,align = 'l',) %>%
  kable_styling(full_width = F) %>%
  column_spec(1, width = "10em", background = "white", include_thead = TRUE) %>%
  column_spec(2, width = "15em", background = "white", include_thead = TRUE)
```
Every function has a `labels` argument to override the default, should it be necessary. Also note when updating labels/units, it is not required to specify all labels/units, only the labels/units that are being updated need be specified.

<div style="margin-bottom:40px;">
</div>

### Global label/units update

We can update labels and units at the global level by calling the desired option within the `options` function. Units for concentration (dv, pred, ipred), time, and tad can also be set using global options and will be appended to the label:

```{r fig.width=9, fig.height=6, out.width="90%"}
options(gof.label.cwres = 'CWRES',
        gof.label.dv = 'MyDrug',
        gof.units.dv="ng/mL",
        gof.units.time="h",
        gof.units.tad="h")
gof(dat)
```

<div style="margin-bottom:40px;">
</div>

### Update labels for a single plot
We can update a default label for a single plot by calling the `label` option as follows.

```{r fig.width=9, fig.height=6, out.width="90%"}
gof(dat, labels = gof_labels(dv='Obs'))
```

<div style="margin-bottom:40px;">
</div>
Finally, the usual `labs()` function can be used to set or modify the labels of any plot. In this case, units do not get applied and must be explicitly included in the labels.

```{r fig.width=3, fig.height=3, out.width="30%"}
gof(dat, 6) + labs(x="Something silly", y="Why not? (light years)")
```


Now let's restore our defaults before moving on...

```{r }
# Restore defaults
options(gof.label.dv    = "Observed Value",
        gof.label.pred  = "Population Prediction",
        gof.label.ipred = "Individual Prediction",
        gof.label.cwres = "Conditional Weighted Residual",
        gof.label.iwres = "Individual Weighted Residual",
        gof.label.time  = "Time",
        gof.label.tad   = "Time After Dose",
        gof.units.dv=NA,
        gof.units.time=NA,
        gof.units.tad=NA
)
```

## Outliers

To highlight outliers using a special symbol (and color), you first define
a column in the dataset which is a factor with 2 levels (be mindful of the
labels, as these will appear in the legend and also note the use of unicode
language for special symbols). The sample data comes from a simulation, and
doesn't have any outliers, but for demonstration purposes we will pretend that
`|CWRES| > 2.5` is a criterion for defining outliers:

```{r}
dat$outlier <- factor(abs(dat$cwres) > 2.5, labels=c("|CWRES| \U{2264} 2.5", "|CWRES| > 2.5"))
```

<div style="margin-bottom:40px;">
</div>

Then, use the `highlight` option, like this:

```{r message=F, fig.width=4, fig.height=4, out.width="50%"}

gof_cwres_vs_pred(dat, highlight=outlier) +
  scale_y_continuous(breaks = c(-2.5,0,2.5))
```

This feature should be used carefully. Note that highlighting values also
populates a legend at the bottom, left corner of the plot area. It works for
multiple panels too:

```{r fig.width=9, fig.height=7, out.width="90%"}
gof(dat, highlight=outlier)
```

## Faceting

Faceting can be applied with `facet_wrap` or `facet_grid` as with any `ggplot` object.  For this example, we will augment our data with some covariate information, specifically sex and study information. In the same folder as the dataset `nmtable.csv` is a file `pkpop.csv` that contains covariate information on PK population.

```{r}
file <- system.file(package="ggcertara", "sample-data/pkpop.csv")

pkpop <- read.csv(file)
head(pkpop)
```

Note that there is one row for each `id` in the dataset. We need to link this
to our goodness-of-fit data by matching on the `id` column. There are many ways
to do this (`merge()`, `left_join()`, ...), but a simple way is to use the
basic `match()` function:

```{r}
i <- match(dat$id, pkpop$id)
dat$sex   <- pkpop$sex[i]
dat$study <- pkpop$study[i]
```

We have now added sex and study to our dataset, but in order to get nice, informative labels, we should convert these to `factor`:


```{r fig.width=9, fig.height=6, out.width="90%"}
dat$sex = factor(dat$sex, levels=0:1, labels=c('Female', 'Male'))
dat$study = factor(dat$study, levels=1:3, labels=c('Study A', 'Study B', 'Study C'))
```

Now, we can produce a plot of DV vs. IPRED faceted by sex as follows:

```{r message=F, fig.width=6, fig.height=3, out.width="60%"}
gof(dat, panels = 3) + facet_wrap(~sex)
```

## Changing aesthetics (colors, etc.)

### Global changes

Changes that will affect all plots can be made with the `update_geom_defaults()` function.  For example, to change the alpha level used for points in all panels:

```{r fig.width=9, fig.height=6, out.width="90%"}
update_geom_defaults("point_c", list(alpha=0.05)) # update the global options
gof(dat)
update_geom_defaults("point_c", list(alpha=0.4)) # reset to default alpha
```

Note that a special custom geom, named `geom_point_c()`, is defined and used in
this package, so that other `ggplot` code will not be affected.


### Changes to individual plots

For a single panel, you can just modify the `ggplot` object.  For example,
let's change the color and symbol of points in a DV vs. PRED plot to be
different by study (which we already added to our dataset earlier):

```{r message=F, fig.width=4, fig.height=4, out.width="40%"}
update_geom_defaults("point_c", list(alpha=0.5, size=2.5)) # make points bigger and more opaque
gof(dat, panels=4) +
    aes(color=study, shape=study) +
    labs(color=NULL, shape=NULL) +
    scale_color_certara()
```

For a `patchwork` object (multiple panels), you can do the same thing by
replacing the `+` operator with a `&`:

```{r message=F, fig.width=6, fig.height=4, out.width="60%"}
gof(dat, panels=3:4) &
    aes(color=study, shape=study) &
    labs(color=NULL, shape=NULL) &
    scale_color_certara()
update_geom_defaults("point_c", list(alpha=0.4, size=1.5)) # restore defaults
```

`patchwork` is smart enough to produce a single legend for both panels (note:
the `theme_certara()` theme must be set globally, as it was at the beginning of
this script, for this to display properly).

## Adding geoms and annotations

As with any ggplot object, geoms can be added. Remember to use the `&` operator instead of `+` when creating multiple panels (this notation comes from the `patchwork` package).

For examples, we can add a dashed line with corresponding label showing the LLOQ to our DV vs. IPRED and PRED plots as follows:

```{r fig.width=6, fig.height=3, out.width="60%"}
 gof(dat, panels = -(3:4)) &
  geom_hline(yintercept=1.2, linetype="dashed") &
  geom_text(data=data.frame(x=Inf, y=1.2), aes(x=x, y=y),
    label="LLOQ = 1.2 ng/mL", hjust=1.1, vjust=-0.5, size=3)
```

As another example, we can add a blue linear regression line in addition to the red LOESS smooth to our panels:

```{r message=F,fig.width=6, fig.height=3, out.width="60%"}
gof(dat, panels=8:9) &
  geom_smooth(method="lm", color=certara_pal()()[1], se=F, size=1)
```

# R session information

```{r}
sessionInfo()
```



